# Heterogeneous Treatment Effects

When we consider heterogeneous treatment effects, we acknowledge that it is very unlikely that every single person reacts to treatment in exactly the same way. But, are there certain *types* of people who always react more strongly to treatment? Are there certain *types* of people who always react less strongly? 

How should we go about (a) looking for HTEs; (b) testings for HTEs with nominal p-value coverages; and (c) reporting HTEs to individuals in a way that make sense? 

## Learning Objectives 

At the end of this week, students will be able to 

1. Understand what an HTE is, and what it is not. 
2. Conduct, test, and interpret models with interaction terms as specific tests of hypotheses. 

## Reading and Discussion: Goodson

1. Why do we call them A/B tests, rather than experiments? Are you uncomfortable with the idea that companies are experimenting on you? Are you uncomfortable experimenting in your subjects? If there is a gap between your feeling about being experimented **on** compared to how you feel when you are **doing** the experimenting, why does this gap exist? 
2. What is an A/B test in Goodson's estimation? 
3. How should we know when a test should be determined to be complete? How should we determine that one experience is *causing* different behaviors in our reference population than another experience? 
  - Can we decide this while we are working through the experiment? 
  - Can we make this choice when we see the outcome that we thought we were going to see? 
  - Can we be *sure* that we have set the correct stopping rules ahead of time? 
4. What are the consequences of peaking, and then stopping early, once we have seen the results? 

## Coding and Demo: The Californians 

```{r, eval = FALSE, echo = TRUE}
library(data.table)
library(stargazer)

makeData <- TRUE

set.seed(1)

if(makeData) {
    sim_size <- 1000
    d <- data.table(californian = sample(c(0,1), sim_size, replace = TRUE),
                    affluence = sample(1:7, sim_size, replace = TRUE),
                    preY = rnorm(sim_size, mean = 100, sd = 7),
                    treat = sample(c(0,1), sim_size, replace = TRUE) )
    d[ , literacy := rnorm(sim_size, mean = d$affluence, sd = 2)]
    d[ , tau      := rnorm(sim_size, mean = 5 + californian)] ## + d$affluence)
    d[ , postY    :=  preY + tau*treat]
}

## What is the (unobserved) true average treatment effect?

## Estimate a model that includes only the treatment effect.
## Interpret all the coefficients. 
mod0 <- 

## Estimate a model that includes the treatment effect and
## the (highly predictive!) pre-treatment Y val. 
mod1 <- 

## Print the two next to each other, and compare what is going on.
## Tell me what is happening in the differences in the intercept
## and the preY coefficients. Tell me what is happening in the SE
## for the treatment effect. 
stargazer(mod0, mod1, type = "text")

## Subset the data into two groups based on gender and estimate
## a model that only includes the preY and treatment effects.
## Print these two models side by side, and tell me what is going
## on.

mod2 <- 
mod3 <- 
stargazer() 

## Based on this, would you conclude that these are different?
## Talk about the 95% CI for each.
## ?confint

## Now, estimate two more models:
##  1. A model with preY, treatment and gender indicator
##  2. A model with prey, treatment, gender, and treatment*gender
##     interaction
##  3. Test for the necessisity of the interaction; first using a
##     t-test. This should be really, really simple. Then, using a
##     f-test for the nested models (see the anova(...) call).
##     Is the p-value the same or different for this test?
##     Why? 

mod4 <- 
mod5 <- 

## ? anova

## Finally, use the results from model 5 to tell me what the treatment
## effect is for males and for californians.


## 
## AT HOME:
## Work to examine what including the other affluence and literacy
## triggers does to your estimates.
##

```


## Coding and Discussion: Tips at a Restaurant 

```{r, eval = FALSE, echo = TRUE}

## Green and Gerber: Question 9.6
## a, b, and c. 

d <- fread('http://hdl.handle.net/10079/cd6be01a-a827-4312-a2fa-74329ce7f96d')

## a. (Probably skip this one)
##    Suppose that you ignored the gender of the server and simply analyzed whether
##    the happyface treatment has heterogeneous effects. Use randomization inference 
##    to test whether the Variance of \tau = 0 using randomization inference by 
##    comparing the variance of potential outcomes in treatment and control. 

## b. Write down a regression model that depicts the effect of the gender of 
##    the waitstaff, whether they put a happyface on the bill, and the interaction 
##    of these factors. 
## 
##    [Write the model here]

## c. Estimate the regression model that you wrote down in (b) and test the 
##    interaction between waitstaff and the happyface treatment. 
##    Is the interaction significant. 

## d. Waiting tables in the time of covid: Suppose that you're on the waitstaff 
##    at this restaurant, and while you're waiting tables you're FULLY garbed 
##    up: facemask, face-shield, full operating gown, and so on. 
##    What this means is that you have the choice to reveal a gender identity
##    that is either "Male" or "Female" to the patrons. 
## 
##    - Is there one gender identity that receives higher tips in this restaurant? 
##    - Is there a gender that has a higher treatment effect? What is the 
##      test that you would run to assess this? 


```

