# Treatment Noncompliance

```{r load packages for unit 08, echo=FALSE, message=FALSE, warning=FALSE}
library(data.table)
library(magrittr) 

library(AER)
```


This begins a section of the class where we are going to evaluate what happens when problems creep into the actual experiments that we are conducting. We are first going to look at what happens when we instruct people to take treatment, but they choose not to. Or, when we instruct people to take control, but they choose to take treatment instead.

It might seem, at first, like we should just proceed by analyzing according to the treatment condition that they actually received. However, because we haven't experimentally assigned this condition, this creates an unprincipled estimator.

This doesn't mean that all is lost however. We can redefine the causal quantity that we are estimating, and produce a reliable estimate of this new concept.We're going to present two such concepts this week. The first concept is the idea of the intent to treat effect (the ITT). The second concept is the idea of the treatment effect among compliers, which we will call the CACE.

## Learning Objectives 

1. **Recognize** when experimental units have not complied with the treatment assignments they were given, and **appreciate** that this causes problems for our two-group estimator.
2. **Recover** causal estimators, but for sub-populations of the overall population. 
3. **Utilize** a new class of model, the two-stage least squares model, or 2SLS, which is the appropriate model choice when we are dealing with either one- or two-sided non-compliance. 

## Starting conversation 

Life on campus is exciting! Whether students are involved in affinity groups, advocacy groups, protest groups, or just party groups, student life on campus is exciting. We're not to be left out! We're not to be denied the chance to make our voices heard. 

But, because we're usually calling in via Zoom, we don't have the typical problems that students and faculty complain about -- parking, enrollment, sports tickets, and beer availability. What we do have to complain about is that "*god awful*" sound of the bell-tower ringing every hour on the hour. 

Suppose that we are to discuss this *very.* *important.* *issue.* before a panel of the deans and University administration. And, further suppose that in light of global events of the past several years, they're actually amenable to what we're proposing -- cutting off those bells, and providing the Berkeley Carillon player a generous retirement. 

![University Carillonist](https://www.youtube.com/watch?v=K_8vta9XDpc)

However, there's a catch. They are worried that taking such an action would be detrimental to the student experience on campus, and they would like to measure the causal effect of playing vs. not playing the Carillon while students are changing classes. 

In breakout rooms, please design an experiment that would be able to measure the difference in student experience. You will have to propose a measurement, a timescale for that measurement, and a feasible randomization that *could actually* occur given the real-world constraints that what is at question are loud sounds emanating from a 300 foot tower in the middle of a large, busy campus. 

If there are any limitations to what you design, please voice those concerns and talk about why they arise, relative to an *ideal* experiment (that you are probably unable to conduct). 
    
## Non-compliance Discussion
### Setting Terms of Understanding 
- What is does the concept of the *intent to treat* effect mean? When is this ITT measurable? When is the ITT an interesting quantity to estimate? When is it uninteresting?  
- What is the compliance with treatment assignment? How does someone measure $ITT_{D}$? Why do Green and Gerber choose to pick such arcane notation? Why is it necessary to know the compliance rate?  
- What is the compliers average causal effect (CACE)? Under what conditions is this CACE guaranteed to be *exactly* the same as the ITT? Under what conditions is this CACE larger than the ITT? Under what conditions is it smaller than the ITT? 
- Is it possible to estimate the CACE without knowing *specifically* who complied? Is it possible to estimate the CACE without knowing anything about compliance rates?   

### Where Does Noncompliance Occur 

Is all of this concern about non-compliance actually a concern? Or, is this just another example of academics getting ahead of themselves and worrying about things that are not actually a concern? 

:::{.breakout} 
In three distinct breakout groups, please talk about one of the following scenarios. 

1. **You are a MIDS instructor writing online content to cause students to be their best possible data scientist.** In this role, you write curriculum, record lectures and assign readings, and create homework assignments for students to work on. How might compliance issues affect curriculum choices? 
2. **You are a non-profit interested in reducing litter at your local surf spot.** In this role, you take steps to raise awareness through signs. How does compliance affect what you are likely to learn in any trial or evaluation of your work? How would you know if someone complies? 
3. **You are a publisher seeking to sell more copies of the newest, and best causal estimators textbook.** You propose to use pre-roll advertisements on video streaming services to get the word out about the new book. What are all the ways that you can imagine measuring compliance? Which would you propose to use, and why?  How might compliance issues affect what you're able to estimate?
::: 

When we come back, each student group will have 5 minutes to talk the other two groups through their scenario, including major risks, opportunities, and learning that they took away from the scenario. 

## Estimating with Non-compliance 

### Estimating with non-compliance 

```{r make data for non-compliance}
## install.packages("AER")      # this has a nice wrapper for iv regression
                                # but we can do it by hand with VERY little work 

nrows = 1000

d <- data.table(
  id  = 1:nrows, 
  y0  = rnorm(nrows, mean=10), 
  tau = rnorm(nrows, mean=5)
)

## create a treatment effect
d[ , y1 := y0 + tau]

## create an assignment vector and a treatment vector with everybody initially
## set to be untreated. 
d[ , assigned := sample(rep(c(0,1), each=nrows/2))] # z in the book
d[ , treated := 0] 

## then, among the people who are assigned to receive treatment, actually 
## treat some of them at random. 
d[
  assigned == 1, 
  treated  := sample(
    x=1:0, size=.N, replace=TRUE, prob=c(.7, .3)
    )]

## finally, observe: 
##   - the potential outcomes to treatment for the treatment group; and, 
##   - the potential outcomes to control for the control group. 
d[treated == 1, Y := y1]
d[treated == 0, Y := y0]
```

With this data created, we can confirm that the different between potential outcomes to treatment and control still produces the treatment effect. If it doesn't we've got bigger problems than compliance! 

```{r confirm tau}
d[ , mean(y1 - y0)]
```

But, the data that we have created to this point has all the data in the science table. In real life, we won't get access to all this data; instead, we get access to observing some potential outcomes for one group and some other potential outcomes for another group. Let's create this restricted set of data to ensure that our estimator can recover the population parameter that we're looking for, even though it only has access to a sample of data from that population. 

```{r create observed data}
d_observed <- d[ , .(assigned, treated, Y)]
```

Does our estimator for the $ATE$ produce an unbiased estimator of the population parameter, given this sample of data? We know that the population parameter has a treatment effect of $5$.

```{r sample estimator }
## how would you code a simple two-group ATE estimator? 

```

### Build Estimators 

Suppose that you cannot /actually/ observe whether someone was treated or not. This will require that you suspend reality for a moment, to suppose that we do not have access to the `treateed` variable. 

In this case, what concept **are** you able to estimate? What concept are you **not** able to estimate? 

#### Estimate the Intent to Treat Effect

Use a linear model to estimate the intent to treat effect. What variables do you need to produce this estimate, and what subset of the data (up to, and including the full set of data) do you use to produce this estimate? 

- What should be the magnitude of your ITT estimator, relative to the actual population average treatment effect that we encoded (i.e. 5)? Why do you anticipate that it will be at this level? 

```{r estimate the itt}

```

### Estimate the Compliance Rate 

Using the full set of data, estimate the compliance rate in two different ways: 

1. Estimate using a `.N` counter, or `mean` or some other such simple transformation on the data.table. 
2. Estimate using a linear model, interpreting the coefficient of that model appropriately. 

What are the trade offs to these two different methods? 

```{r estimate the compliance rate two ways. }

```

### Compute the CACE 

Because you have the ITT and the compliance rate, estimate the CACE. Once again, compute this compliance rate in several ways. 

1. Using a `.N` counter, or `mean` or some other such simple transformation on the **entire**  data.table. Once again, you might notice that this does not have a sampling based uncertainty estimate associated with it. You need not code this now, but talk about how you would produce something akin to a standard error for the mean given this data. 
2. Using a `.N` counter, or `mean` or some other such simple transformation on a reasonable **subset** of the data.table. 
  - Will the estimate that you produce on this subset of data be larger, smaller, or about the same as the estimate that you produced on the full data? 
  - Will the estimate of the standard error for the mean be larger, smaller, or about the same as the estimate that you produced on the full data? 
3. Finally, estimate the CACE using two linear models. How, if at all, would you produce an estimate of the sampling based uncertainty of these estimates? 

```{r compute cace several ways}
cace_one   <- 'fill this in' 
cace_two   <- 'fill this in' 
cace_three <- 'fill this in' 
```

How do you feel about the sampling based uncertainty that you can produce with these estimators? 

## Two Stage Least Squares

In order to estimate with a reliable standard error, we can turn to two stage least squares.

Two-stage least squares estimators have the benefits of

1. Doing _exactly_ the same thing that the $CACE = ITT / ITT_{d}$; but,
2. Doing it in a way that has known standard errors that are quickly and easily computable. 

### First Stage 

In the first stage we: 

- Estimate the proportion of people who are receive treatment as a function of being assigned to treatment. 
- In the case of one-sided non compliance this is _exactly_ the same thing as estimating the
   ITT_{d}, right?
   
```{r 2sls first stage}
first <- 'fill this in'
```
   
```{r the make predictions}
## calculate the fitted values from this regression
##   - that is, just multiply the coefficients that you estimate from the
##     first stage times the data values. In the event that the exclusion
##     restriction holds, then these predictions are just orthagonal to every
##     thing that is not modeled in your data! 

```

### Second stage 

In the second stage we: 

- Estimate the relationship between the predicted values and the outcome. 
- This will just tell you how the outcome changes in response to the amount of change that your treatment assignment is able to produce.


```{r 2sls second stage}
second <- 'fill this in'
# coeftest(second, vcovHC(second, type = "const")) ## these ses might be wrong. 
```


```{r AER}
## I'll note that the standard errors from this "hand-rolled" 2SLS will not
## be correct (due to some accounting issues in the variance between the predictions
## in the first stage and the second stage.
##
## we can fix this by hand -- though I wouldn't -- or we can use a library that will
## do the accounting for us, from the library AER

# library(AER)
# iv.model <- ivreg(Y ~ treated | assigned, data = d2)
# 
# coeftest(iv.model, vcov = vcovHC(iv.model, type = "const"))

```